### 2016-07-22 Projects/DelCo_2012/Report/DelCo_2012.tex
Complete draft of text

All of the fire conditions generated during these experiments were ventilation limited or ventilation controlled fires. In other words if more oxygen was supplied to the fuel gases available in the fire compartment and more products of combustion were exhausted from the fire compartment, the heat release rate of the fire would have increased. There are simple formulas which may provide some insight here, although for full disclosure these simple algorithms have not been validated with the geometry that we used in these experiments.

If we have a single open doorway to a residential scale room, which is approximately 0.9~m (3 ft) wide and 2.1~m (6.9 ft) high, the minimum heat release rate needed to flashover the room would be about 2 MW and the amount of heat release rate that the doorway could support assuming complete combustion of the oxygen that would flow through the doorway opening via natural convection would be approximately 4 MW.  If we double the width of the open doorway (the doorway to the lower level of the two story structure)  Then the heat release rate that could be generated inside the compartment, this does not account for flames burning outside of the compartment, would be approximately 8 MW.

The flowrate used for these experiments were sufficient to quickly cool the hot gas layer and suppress each of the fires.

static nozzle lines all performed the same.

movement, sweeping nozzle, etc -- more efficient cooling / more effective that the type of nozzle or the amount of suppressive agent.
air entrainment - ref hose stream study.

@@

Experiments were conducted to investigate the capabilities and limitations of compressed air foam (CAF) relative to water when used for interior structural firefighting.  The experimental designs were developed based on a project planning workshop led by the Fire Protection Research Foundation.  The key outcomes were to measure the gas cooling and fire suppression capabilities of CAF compared to water. Incidents where firefighters had experienced burn injuries or other challenges while using CAF during interior attack operations in residential occupancies served as the catalyst for this study.

Full scale fire experiments designed to address those questions were conducted over a period of years from 2012 through 2015. The experimental arrangements were designed to be representative of residential fire scenarios.

Gas Cooling

Fire Suppression

>>>>

Shortly after the NFA equation was issued, another Federal agency, The National Bureau of Standards (now NIST) released a finding about the relationship of oxygen to the heat release rate of a fire.  They found that for many fuels, approximately 13.1~MJ of heat were generated for every kilogram of oxygen comsumed during the combustion.  This relationship then led to simple algorithms that addressed the relationship of the ventilation available o the room, in other words how much oxygen can flow through an opening, with the theoretical heat release rate which assumes stoichiometric conditions or 100~\% effecient burning where all of the fuel and the oxygen is consumed.

All of the fire conditions generated during these experiments were ventilation limited or ventilation controlled fires. In other words the conditions in the compart were fuel rich.  If the ventilation was increased to allow more oxygen to be supplied to the fuel gases available in the fire compartment and more products of combustion were exhausted from the fire compartment, the heat release rate of the fire would have increased. There are simple formulas which may provide some insight here, although for full disclosure this simple algorithm has not been validated with the geometry that we used in these experiments.  The purpose here is to get a sense of the amount of heat that the hose streams need to overcome.  This algorithm, just like the NFA formula has never been validated for this application.

\begin{equation} \label{eq:HRRSto_item}
\dot{Q}_{stoichiometric} = 1500 Area \sqrt{Height} \\
\end{equation}

In this equation, the maximum heat release rate, in~kW, based on airflow through the area of the opening, in~m^2, represented by Area. The other input needed is the height of the opening in~m, shown as Height.

If we have a single open doorway to a residential scale room, which is approximately 0.9~m (3~ft) wide and 2.1~m (6.9~ft) high, the amount of heat release rate that the doorway could support assuming complete combustion of the oxygen that would flow through the doorway opening via natural convection would be approximately 4 MW.  If we double the width of the open doorway (the doorway to the lower level of the two story structure)  Then the heat release rate that could be generated inside the compartment would be approximately 8 MW, this does not account for flames burning outside of the compartment.  With the single vent opening to the outside, the relationship should hold.  In the single story, we had an additional window opening in some of the experiments which added ventilation to the room.  In the two story we had an another open doorway that allowed exhaust gases from the fire compartment to flow up the stairs, through the first floor and out of the structure.  Even with the additional ventilation in both cases, the conditions in the fire compartment remained fuel rich as the flames extended out of the openings from the structures. So it would seem that inside the compartment we would not have exceeded the 4~MW for the one story or the 8~MW for the two story stucture. Eariler we had noted that the latent heat of vaporization of water is 2255~kJ/kg or approximately 2.3~MJ/kg.  Assuming ideal heat transfer the mass flow rate of water needed to absorb the heat being generated inside the fire compartment would be approximately 1.8~kg/s for the one story and 3.5~kg/s for the two story.  In our experiments our mass flow rate of water was 7.57~kg/s (16.7~lb/s), which ideally is twice what was needed to absorb the heat in the fire compartment of the larger fire in the lower level of the two story structure. Again this is a simplified analysis, much like the current tools provided in fire department literature for the required fire flow rate.  However, it demonstrates that for a residential scale room, in this study upto to 51.4~$m^2$ (553~$ft^2$) with the equivalent of two open doorways supplying fresh air, a flow of 2~gallons per second or 120~gpm of water should be adequate quickly suppress the fire in the compartment.

In all of the suppression experiments, the fire was controlled with total flow of 227~L (60~gal) or less of agent. During the course of the suppression experiments the fuel packages were modified to incorporate more wood based fuels.  The thinking of some of the technical panel members was that the synthetic fuels, such as polyurethane foam, burned mostly at the surface and therefore they were easily cooled and suppressed.  The wood based fuels, develop a char layer and therefore the burning would be "deep seated", which might provide a more significant suppression challenge.  Since the Class A foam solution has a reduced surface tension, which could allow the solution to penetrate and cool the burning wood based fuels better, it was thought that with more exposed, burning wood surface, the better the opportunity for CAF to demnostrate a difference.  As shown in Chapter 2, the wood fuel load constisted of stacks of wood pallets, plywood lined walls, and an exposed wood "ceiling" that was constructed of wood "i-beams" and oriented strand board.  The experiments conducted with the wood fuel load did not demonstrate a significant difference between water and CAF.

The final set of suppression experiments used a larger space than the previous two sets of suppression experiments in the single story structure.  Again the thought was to create a bigger fire challenge.  As discussed previously, with the larger area and the double the ventilation, the fire should be more of a challenge to the hose streams.  There was also another variable that would be introduced, a firefighter controlling the nozzle.  In the suppression experiments conducted in the single story, the nozzles were in a fixed position.  As a result, after the initial fire knockdown, it was typical that the agent was not able to hit all surface of the burning fuel sufficiently and a rekindle would occur.  This can be seen for both the water and CAF experiments in the data provided in Appendix B.  In some single story experiments, the fire was given time to redevelop and additional nozzle flows from the fixed positions were conducted to collect more data and to see if we could extinguish the fires. Would the CAF build and flow to burning areas that were not previously impacted by the hose streams. During the additional flows, no significant differences were seen.  In all of the single story experiments firefighters with a hoseline were needed to extinguish the remaining fire and hot spots.  So for the last set of experiments, the choice was made to use a firefighter so that the nozzle would be moved in an intelligent manner to make the best use of the hose stream and suppress the fire in the lower level of the two story.

The same firefighter was used for all of the experiments conducted in the two story.  The firefighter was from the MCFRS and had been trained in  in the use of CAF and had experience using CAF.  The firefighter used the reach of the hosestream to cool the fire gases just inside the doorway as a means to knockdown the flames that were extending up to the upper level of the test structure.  Then the hoseline was advanced and the nozzle was moved rapidly to hit the fuels inside the compartment.  The nozzle was shut down when there were no flames visible from the doorway.

The flow times are all less than 30 seconds. So even though the fire size had increased, a total flow of 227~L (60~gal) or less of agent was used to quickly cool the hot gas layer and suppress each of the fires. This is consistent with the ventilation analysis above.  The nozzle movement clearly had a positive impact on suppressing the fire in the 11~m (36~ft) deep fire compartment.

During the workshop there were discussions that CAF could suppress a fire perhaps as much as five times faster than water. During the course of this study there were no significant differences demonstrated between CAF and water for use in interior structural firefighting on a residential scale.  
Going back to the premise of the study preventing firefighters from being injured during an interior firefighting operation with CAF, there is nothing in the fire suppression results to suggest that CAF would present a thermal hazard, provided that an adequate amount of water is being flowed for the given fire condition.

\Chapter{Future Research}

One of the challenges with suppression research or determining the required flowrate of water needed on a building fire is the lack of data on realistic baseline requirements of the amount of water needed for gas cooling and suppression.  In this study it seems clear that there was a sufficient amount of water for gas cooling and suppression.  So one question would be, how much could the flow rate of water have been reduced before a significant difference in performance was observed?

Information from actual incidents in the field, shaped this study.  From the gas cooling and suppression perspective, the data does not indicate that the use of CAF would result in a higher thermal hazard.  However, there are other issues like nozzle reaction force and flow separation which have been covered in the Cal Poly fire ground studies~\cite{Carracino:2013,Dicus:2013,LaPolla:2012}. These are issues that can be addressed in training.  It is important to provide adequate training when adopting a new method or new technology.  Perhaps these types of issues should be examined further to better understand how suppression activities take place in the field.  What is the relationship between the training and the actual fire suppression activties that occur?

Additional research is needed to examine the impact of nozzle type and nozzle movement on the fire environment.  It is possible that some of the discomfort or injuries that firefighters experience during interior firefighting could be the result of a poor choice of nozzle or nozzle movement that results in displacing hotter gases into the area where the firefighter is located.  Currently the fire service relies on experience to develop a feel for the correct combination of nozzle type and movement.  A study may be able to demonstrate a clear casue and effect relationship between the amount of water flowing, the type of nozzle, and the movement of the nozzle and how that can impact the flow of gases within the structure. 
Based on the results of this study, it would seem that training and understanding the impact of nozzle movement are ways to look for more effective use of water.

@@

Experiments were conducted to investigate the capabilities and limitations of compressed air foam (CAF) relative to water when used for interior structural firefighting in residential stuctures.  The experimental designs were developed based on a project planning workshop led by the Fire Protection Research Foundation.  The key outcomes were to measure the gas cooling and fire suppression capabilities of CAF compared to water. Incidents where firefighters had experienced burn injuries or other challenges while using CAF during interior attack operations in residential occupancies served as the catalyst for this study.

Full scale fire experiments designed to address those questions were conducted over a period of years from 2012 through 2015. The experimental arrangements were designed to be representative of residential fire scenarios. The water and CAF flows were based on those used by a fire department with 35 CAFS front line engines.  This same department also experienced injuries to firefighters while using CAF during interior firefighting operations, in residential stuctures.  This study is composed of three types of experiments; spray density, gas cooling and fire suppression. In each type of experiment a comparison was conducted between plain water and compressed air foam being discharged from nozzles of similar type with similar mass flow rates of water.  The flowrate for each experiment was 120~gpm of water.  For the CAF, 60~cfm of air and a 0.3~\% flow of Class A foam concentrate was added to the water flow.

Thirty-two spray density tests were conducted to determine where the water would land within a compartment based on various nozzle patterns and locations. The gas cooling and fire suppression tests simulated an interior fire attack through the deployment of hose lines inside the structures with fixed nozzles. The purpose of these spray density tests was to determine the appropriate nozzle locations for the gas cooling and fire suppression tests to follow by gaining an understanding of the water distribution throughout the structures.  This report provides a complete set of the spray density results in graphic form.

Eighty-eight gas cooling experiments were conducted to study the effectiveness of cooling the upper gas layer in a space adjacent to the area of fire involvement. These tests examined five different hose stream patterns, two nozzle types, and both compressed air foam and water as the suppression agent. The focus of this experimental series was to study the impact of the test variables on the hot gas layer temperatures when a fixed hose stream was applied to a room adjacent to the fire room.  Suppression agents were not applied to the fire in the burn room.  In this report an aggragate analysis is made to examine difference due to the loaction of the instrumentation relative to the impact area of the agent and compare the differences in the amount of gas cooling between plain water and CAF hose streams.  No significant differences were found in the gas cooling capabilities of the two agents.  A more detailed analysis of the individual experiments was conducted by Mitchell of Cal Poly~\cite{Mitchell:1}.  The conclusion of both analysises are in agreement.  The full gas cooling data set can be found in the Mitchell report.

There were two different geometries used for the fire suppression experiments. Twelve of the suppression experiments were conducted in a single story structure with an entry hallway that led to a fire compartment with a floor area of 27~$m^2$ (291~$ft^2$).  Within this structure, two different types of fuel loads were used.  In one case, most of the fuel consisted of synthetic materials in the upholstered furniture, carpet, and padding. The other fuel load used only wood based fuels, to examine the value of the reduced surface tension of the water in the CAF and its interaction with the surface of the wood fuel. In the experiments conducted in the one story building, a fixed nozzle was flowed from the hallway for approximaetly 15~s and then another fixed nozzle located in the fire comparment was flowed for an additional 15 seconds.

Seven of the suppression experiments were conducted in a two story structure.  The fire compartment was the full lower level of the stucture and it had an open floor plan with an area of 51.4~$m^2$ (553~$ft^2$).  At one end of the room was an open doorway, 1.8~m (6~ft) wide by
2~m (6.7~ft) high. A manual hose stream application was made from outside the doorway on the lower level. In these experiments there was a combiantion of wood based and synthetic fuels. The fires were allowed to transition through flashover and then the initial fire suppression attack was made.  In each case the initial knockdown of the fire required 60~gal of agent or less.
